Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper-prediction, extend equations to hybrids

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes.
For the sake of clarity, we do not distinguish between models for Exp1 and Exp2
at this point.
It should be noted, however, that in the case of Exp2 the term `genotype' refers
strictly to the parent lines.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance:

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\mathbf{W} =
\begin{bmatrix}
  \mathbf{\hat{W}}_{1} \\
  \mathbf{W}_{2}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\mathbf{W}_{2}$ while marker
covariates for genotypes not covered by the incomplete predictor
($\mathbf{\hat{W}}_{1}$) need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} =
\begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix},
 \mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_1\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_2
 \end{bmatrix},
 \mathbf{U} =
\begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix},
\end{equation}

where $\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} 
\mathbf{K}^{-1}_{22} \mathbf{W}_{2}$.

@@

Our final model looks like this:

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}\mathbf{W} \boldsymbol{\alpha}_{l} + 
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{U}_{l} \boldsymbol{\epsilon}_{l} +
\mathbf{e}.
\end{align}

Breeding values or general combining abilities can be estimated as:

@@

For the hybrid data, the obtained GCA effects for the inbred lines represent 
half their breeding values.
The predicted agronomic performance is:

\begin{equation} \label{eq:predicted-performance}
\hat{\mathbf{y}} = \sum_{l=1}^{L} \mathbf{Z}_{l}\hat{\mathbf{g}}_{l} 
\end{equation}

>>>>

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptomic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes.
For the sake of clarity, we do not distinguish between models for Exp1 and Exp2
at this point.
It should be noted, however, that in the case of Exp2 the term `genotype' refers
strictly to the parent lines.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance:

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_{1}\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_{2}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\mathbf{W}_{2}$ while marker
covariates for genotypes not covered by the incomplete predictor
($\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} \mathbf{K}^{-1}_{22} \mathbf{W}_{2}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} =
\begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix},
 \mathbf{U} =
\begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}.
\end{equation}

@@

The above equation implies that ultimately single step prediction uses only
one random effect vector $\mathbf{g}$, which entails the observed effects of
the incomplete predictor $\mathbf{g}_{2}$ and the imputed effects
$\mathbf{\hat{g}}_{1}$ for genotypes that are not equipped with the incomplete
predictor plus the residual imputation error $\boldsymbol{\epsilon}$.

Our final model is now:

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}\mathbf{W} \boldsymbol{\alpha} +
\mathbf{Z}\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e}.
\end{align}

Extending this model to hybrids with parent lines from two heterotic groups
simply requires us to separate the feature matrices of the two predictors
into a paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D}\mathbf{W}_{D} \boldsymbol{\alpha}_{D} + 
\mathbf{Z}_{D}\mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F}\mathbf{W}_{F} \boldsymbol{\alpha}_{F} + 
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}

Breeding values or, separately for each heterotic group, general combining
abilities can be estimated as:

@@

For the hybrid data, the obtained GCA effects for the inbred lines represent 
half their breeding values.
The predicted agronomic performance of the inbred lines is

\begin{equation} \label{eq:inbred-predicted-performance}
\hat{\mathbf{y}} = \mathbf{Z}\mathbf{\hat{g}} 
\end{equation}

and that of the hybrids is equivalently

\begin{equation} \label{eq:inbred-predicted-performance}
\hat{\mathbf{y}} = \mathbf{Z}_{D}\mathbf{\hat{g}}_{D} + \mathbf{Z}_{F}\mathbf{\hat{g}}_{F}.
\end{equation}
